Topic modeling on EU and the USA research projects

### 2017-06-26 thesis_report/draft_thesis.tex
Fixed: mistakes in the report (Introduction-Methodology)

Nowadays, there are many \todo{ That sounds like a joke. Research
  funding is extremely competitive. } funding opportunities for
researchers. For instance, European Union countries are encouraged to
invest 3\% of their GDP in research and development by 2020. EU's
latest research programme, Horizon 2020, allows researchers to
identify new opportunities and directions in almost any field of
research. The programme's goals are to strengthen the EU's position in
science, support industrial innovation and address social concerns,
such as climate change, green transportation and energy.

Another funding agency that actively promotes the progress of science
is the National Science Foundation in the United States of America.
NSF supports research in all the non medical fields of science \&
engineering and aims to advance the national health and prosperity. It
finds \todo{funds} approximately 24\% \todo{Where did you find this
  number? Add reference.} of the federally supported research
conducted by the USA's colleges and universities.

With our work, we aim to investigate the evolution of scientific
research in the EU and the US. In other words, we attempt to identify
which subjects are more likely to get a funding, based on the previous
focus the two regions in research fields \todo{Does not read right}.
In order to answer these questions, we tried to gain some insights
from the projects that got funded the last two decades under the EU
Framework Programmes and the NSF in USA.

When researchers apply for funding, one of the first steps they do is
to identify an interesting subject of the many possible topics of
scientific investigation. The topic(s) of a project is always
described in the project's abstract/description. Obtaining
\todo{Using} these abstracts, one can possibly investigate the topics
of the projects and find the subjects that are being addressed and
gain some information regarding the popularity of the topics among
others.

In this work, we discover topics and their evolution from 1994 to
2017, using the projects' abstracts that got funded \todo{in} this
period. Using a topic model, we extract a set of topics and
investigate the changes in the content per year/framework programme.
These topics are used to gain insight into some of the research that
occurs the last two decades. Our goal is to share these insights with
researchers and encourage them to develop better strategies to
establish themselves as more effective researchers globally, by
choosing a subject that is more likely to get a funding.

In section 2 \todo{Throughout: never use hard-coded references. Always
  use LaTeX cross-references.}, we briefly talk about the papers that
mostly influenced our study. Section 3 introduces some basic concepts
in topic modeling and we describe the model that we used for topic
extraction, Latent Dirichlet Allocation, which is a statistical
generative model for documents. In section 4 we provide information
about the dataset structure and the methodology that we followed in
our analysis. In section 5 we interpret the results of our analysis,
\todo{and we}discuss the evolution of the topics over time and the
differences between the two regions' topics based on their similarity.
Finally, section 6 summarizes our work and section 7 suggests some
future work.

\section{Related work}\todo{Having this section here, before any
  introduction to LDA, makes counter-intuitive reading. Better move it
after.}

This chapter \todo{section} surveys previous work in topic modeling on
scientific data. Topic modeling algorithms provides \todo{provide} a
suite of algorithms to discover hidden thematic structure
\todo{structures} in large collections of texts and can be adapted to
many kinds of data. These algorithms can be used to summarize,
visualize and explore a collection of documents. The papers listed
below are applying a Latent Dirichlet Allocation model in an
interesting approach on their corpus.

D.Blei's \todo{Throughout: separate initials with space from surname}
research focuses on probabilistic topic models, Bayesian methods and
approximate posterior inference. In his work "Introduction to
Probabilistic Topic Models",\cite{Blei11introductionto}
\todo{Throughout: Fix quotation marks; also citations are placed
  before punctuation} he describes the basic ideas behind LDA and
applies the model to articles from the journal "Science"
\todo{\emph{Science}}. The model assumes that the topics are generated
first, before the documents and that the documents exhibit multiple
topics, but each one of them exhibit those topics in different
proportion. In our implementation we also use the LDA model and we
follow a similar approach, since we don't use any previous information
about the subjects that run through our dataset and the documents are
not labeled with keywords or topics. The topic distributions arise by
computing the hidden structure that resembles the thematic structure
of the collection.

The paper that mostly influenced our work is "Finding Scientific Topics", by
T.Griffiths and M.Steyvers.\cite{griffiths_steyvers04} They investigated
abstract data from articles published in the Proceedings of the National Academy
of Sciences (PNAS) from 1991 to 2001 and compared research topics obtained from
topic modeling with existing categories. In our implementation, we also use 
the abstracts of the projects as documents for the model.

Griffiths and Steyvers describe LDA and present a Markov Chain Monte Carlo
algorithm for inference in the model, illustrating the operation of the
suggested algorithm on a small dataset, and then on a larger corpus of PNAS
abstracts. Finally, they define the dynamics of the topics as a means of gaining
insight into the dynamics of science. As "hot" topics are identified the topics
that appear more frequently over the years (positive linear trend), and as
"cold" the ones that appear less frequently (negative linear trends over time).
We also analyze the topics over time, in order to highlight potential trends in
scientific research in EU and USA.

Inspired by Griffiths and Steyvers, L.Sun and Y.Yin wrote a paper,
"Discovering themes and trends in transportation research using topic
modeling", \cite{Sun201749} about topic modeling in transportation
research. Considering approximately 17000 abstracts of articles
published on 22 leading journals in transportation research as
documents, they infer 50 topics by using LDA. Compared to "Finding
scientific topics", Sun and Yin use more than one journal (22 popular
journals) and they investigate the journal-topic distribution and
journal similarity. In our analysis, we identify the EU and the USA as
2 regions (like 2 different journals), with different proportions of
documents-projects, and we visualize the topic similarity of the two
regions per framework programme.

\section{Theoretic background}

In this chapter we will present some basic theoretic background about
topic modeling. To deal with extensive text body \todo{probably
  plural}, machine learning researchers have developed topic modeling:
a suite of methods that analyze the words of the original text to
discover the themes that run through the data and detect hidden
semantic structures by analyzing a corpus. By marking up the documents
using these topics and by using the resulting structure, the model can
be used for organization, finding similar documents, or any other
similar tasks. A topic is here defined as a probability distribution
of words, meaning that certain words are more likely within a certain
topic. Since the topics emerge from the analysis of the original
texts, topic modeling algorithms do not require any prior annotations
or labeling of the documents.\cite{Blei11introductionto}

@@

Some further common concepts and terms in topic modeling: 
\begin{itemize}
\item[] A \textit{word} is defined as an item from a vocabulary
  indexed from ${1, ..., V}$, where V \todo{$V$} is the size of the
  vocabulary. All words are represented as unit-basis vectors with one
  component equal to one and the rest equal to zero.
\item[] A \textit{document} is a collection of words denoted by
  $w = {w_1, ..., w_N}$, where $w_n$ is the $n$th word and $N$ is the
  total number of words in the collection.
\item[] A \textit{corpus} is a collection of documents in a dataset.
  It is denoted $D = {w_1,..., w_M}$, where $w_m$ is the $m$th
  document in the corpus and $M$ is the total number of documents.
\item[] \textit{Latent variables} are variables that may not be directly observed, 
unlike observable variables. Latent variables can instead be inferred 
from other observable variables.
\item[] \textit{Polysemy} is the capacity for a word to have multiple related meanings. 
An example of this is the word “plant” which can mean a living organism of 
the kind exemplified by trees, herbs etc., or a place where an industrial or
manufacturing process takes place.
\item[] \textit{Synonymy} is the capacity for several words to have similar meanings such as
the words “buy” and “purchase”.
\end{itemize}

@@

When introducing the Latent Dirichlet Allocation in section 2.2.3
\todo{Cross reference}
\todo{As you can see this is a disaster: always cross-reference with
  LaTeX}, we will be using the terms prior and posterior probabilities
and likelihood function:

\begin{itemize}
\item The prior probability distribution of and \todo{an} uncertain
  quantity is the probability distribution that express the value of
  this quantity before some evidence is taken into account.
\item The posterior probability is the conditional probability after some 
evidence is taken into account.
\item The likelihood function of a set of parameters given some outcome 
is equal to the probability of observing said outcome given those 
parameter values.
\end{itemize}

@@

In the context of probabilities, the Jensen Inequality is stated as
follows: If X \todo{$X$} is a random variable and f \todo{$f$} a
convex function then \todo{What is $\mathbb{E}$?}:

@@

And, if f \todo{f} is a concave function the inequality turns:

@@

@@

\begin{equation}
D_{KL}(P \mid \mid Q) = \int_{-\infty}^{+\infty} p(x) log\frac{p(x)}{q(x)}dx
\end{equation} \todo{ log in mathematics is $\log$}

This can also be expressed using the definition of expectation (and
expressing the log of a division as difference of log) as:

\begin{equation}
  D_{KL}(P \mid \mid Q) = \mathbb{E}_{p}[log p(x)] - \mathbb{E}_{p}[log q(x)]
\end{equation} \todo{Smae with log; also, it is a function.}

@@

Once we have a representation for individual documents, the natural
next step is to select a model for a set of documents. Given a
training set of documents, we will choose values for the parameters of
a probabilistic model that make the training documents have high
probability. Then, given a test document, we can evaluate its
probability according to the model. The higher this probability is,
the more similar the test document is to the training set. \todo{Is
  this really what you use the multinomial distribution for? You use
  the multinomial distribution because LDA assumes that this is the
  distribution of topics in documents and words in topics.}

The probability distribution that we use is the multinomial. In
probability theory, the multinomial distribution is a generalization
of the binomial distribution. For example, it models the probability
of counts for rolling a k-sided dice n times. For n independent trials
each of which leads to a success for exactly one of k categories, with
each category having a given fixed success probability, the
multinomial distribution gives the probability of any particular
combination of numbers of successes for the various
categories\footnote{\url{https://en.wikipedia.org/wiki/Multinomial_distribution}}.
\todo{Variables like $k$, $n$, etc. should be in math mode.}

@@

The Dirichlet distribution is a prior for the multinomial
distribution. This means that if the prior distribution of the
multinomial parameters is Dirichlet then the posterior distribution is
also a Dirichlet distribution (with parameters different from those of
the prior). \todo{ If the Dirichlet is the prior for the multinomial
  then the posterior is the multinomial; in fact the Dirichlet is the
  conjugate prior of the multinomial. It is the posterior distribution
  of the \emph{parameters} that is still Dirichlet. }

The Dirichlet process is a way to model randomness of a probability
mass function (PMF) with unlimited options (e.g. an unlimited amount
of dice in a bag). The process is similar Polya's
Urn\footnote{\url{http://www.statisticshowto.com/dirichlet-distribution/}},
only instead of having a set number of ball colors you have an
unlimited amount: \todo{Why do we need to know that? }

\begin{itemize}
\item Start out with an empty urn.
\item Randomly pick a colored ball and place it in the urn.
\item Then choose one option:
\begin{enumerate}
\item Randomly pick a colored ball and place it in the urn.
\item Randomly remove a colored ball from the urn, then put it back with another 
ball of the same color.
\end{enumerate}
\end{itemize}

As the number of balls in the urn increase, the probability of picking a new
color decreases. The proportion of balls in the urn after an infinite amount of
draws is a Dirichlet process.

@@

First, we'll introduce some basic models for topic modeling. The
Latent Semantic Indexing, or LSI, was presented by Deerwester et al.
in 1990. \todo{Add reference.} The model manages to deal with the
problem that multiple terms can refer to the same meaning, i.e.
synonymy. However, it is not as successful regarding polysemy. The
reason is that every term is represented as just one point in the
so-called latent semantic space. Furthermore, a word that can mean two
or more different things is represented as a weighted average of the
different meanings.

After LSI was introduced, Hofmann presented the Probabilistic Latent
Semantic Indexing, or PLSI model \todo{Add reference}. PLSI is a topic
model where each word in a document is generated from a single topic
which results in that each document in a corpus can be represented
with a topic distribution. Later, in 2003 Blei et al. presented the
Latent Dirichlet Allocation, or LDA \todo{Add reference}. As opposed
to PLSI, LDA is a statistically generative model for documents where
each word in a document can be generated by all topics.

@@

Latent Semantic Indexing was one of the earliest methods for finding
relationships between documents and the words that occur in
them.\cite{Deerwester90indexingby} In LSI a term-document matrix is
created by analyzing the corpus, where the rows correspond to words
and columns to documents. Each element in this sparse matrix describes
the number of times a word occurs in a document but this term count
can also be weighted with for instance TF-IDF \todo{Explain}. Letting
each word represent a dimension in a very high dimensional space, a
document can be seen as a vector with components corresponding to its
weighted term counts. \cite{Salton:1988:TAA:54259.54260}

@@

\subsubsection{Latent Dirichlet Allocation (LDA)}

LDA is a topic model that was first presented as a graphical model for topic
discovery by David Blei, Andrew Ng, and Michael I. Jordan in 2003. The basic
idea is that documents exhibit multiple topics. LDA is a generative statistical
model that allows sets of observations to be explained by unobserved groups
that explain why some parts of the data are similar. For example, if
observations are words collected into documents, it posits that each document
is a mixture of a small number of topics and that each word's occurrence is
attributable to one of the document's topics.\cite{Blei11introductionto}

This is similar to PLSI, except
that in LDA the topic distribution is assumed to have a Dirichlet prior. 
In practice, this results in more reasonable mixtures of topics in a
document. The reduced document description result in PLSI is a vector where
each element describes the mixing proportion for a topic. A limitation with
PLSI is that there is no generative model for these proportions, making it
difficult to handle unseen documents. The Latent Dirichlet Allocation model
tries to solve this limitation by setting a Dirichlet prior on the topic
distribution.\cite{blei2003latent} \todo{In essence you repeat what
  you said in the beginning of the paragraph.}

@@

The basic idea in LDA is that we define a topic to be a Dirichlet
distribution over a fixed vocabulary. \todo{No, a topic is a
  multinomial distribution.} Technically, the model assumes that the
topics are generated first, before the
documents.\cite{Blei11introductionto} Now for each document in the
collection, we generate the words in a two-stage process:

@@

This statistical model reflects the intuition that documents exhibit
multiple topics. Each document exhibits the topics with different
proportion (step 1); each word in each document is drawn from one of
the topics (step 2b), where the selected topic is chosen from the
per-document distribution over topics (step 2a). All the documents in
the corpus share the same set of topics, but each document exhibits
those topics with different proportion - that's the distinguishing
characteristic of LDA. A more detailed explanation of the model
follows in section 2.2.3.1. \todo{Again, fix cross-references}

\paragraph{Model}

LDA is a probabilistic model of text documents, which assumes a
collection of K topics. Each topic defines a multinomial distribution
over the vocabulary and is assumed to have been drawn from a Dirichlet
$\beta_k \sim Dirichlet(\eta)$. \todo{What is $beta_k$ and what is
  $\eta$? If you explain them later (which you do) this sentence
  should follow the explanations.}

In LDA, each document is represented by a mixture of the topics, with
weight $\theta_w^2$ \todo{Where does the square come from?} for topic
$z$ in document $w$. These weight \todo{weights} have a Dirichlet
prior, parametrized by a hyper-parameter $\alpha$. Each topic is a
probability distribution over the vocabulary, with probability
$\beta^w_z$ for word $w$ in topic $z$. We define the following
variables and notation:
\begin{itemize}
  \item[] $k$ is the number of topics.
  \item[] $V$ is the number of unique words in the vocabulary.
  \item[] $\theta$  is the topic distribution (of length $k$) for a document,
   drawn from a uniform Dirichlet distribution with parameter $\alpha$.
 \item[] $z_{n}$ is a topic assignment for $word w_{n}$ \todo{word
     $w_n$}, sampled from $p(z_{n} = i|\theta) = \theta_{i}$.
  \item[] \textbf{w}$ = (w_{1}, ... , w_{N})$ is a document with $N$ words.
  \item[] $w_{n}^{i} = 1$ means that the word $w_{n}$ is the $i$-th word 
  of the vocabulary.
  \item[] $\beta$  is a $k \times V$ matrix, where each row $\beta_{i}$ 
  is the multinomial distribution for the $i$-th topic. That is, 
  $\beta_{ij} = p(w^{j} = 1 | z_{j} = i)$.
\end{itemize}

@@

We can see that this does in fact generate a document based on the topic 
mixture $\theta$, the topic-word assignments $z$, and the probability 
matrix $\beta$. Then, we can analyze a corpus of documents with LDA by 
examining $\beta$, which is the posterior distribution of topics, $\theta$ 
and $z$ conditioned on the documents. This reveals latent structure in
the collection that can be used for prediction or data exploration.

@@

This
posterior can't be computed directly; it is usually approximated using Markov
Chain Monte Carlo (MCMC) or variational inference. Both methods are effective,
but also present significant computational challenges in large data sets
~\cite{onlineLDAvb}. In our implementation, we're using the online variational
LDA model, which is described in section 2.2.3.4. \todo{Cross reference.}

The key difference in LDA compared to PLSI is that the process outlined above
actually is generative for documents. This means that, given a new document,
one can use the parameters learned previously to estimate the topic
distribution $\theta$ for this new document. \todo{You have already
  said that many times.}

A limitation with the standard LDA model is the fix \todo{fixed}
vocabulary. This means that new words cannot enter the model and that
you have to know a priori what words will model the corpus
effectively. A setting where this could be problematic is when
analyzing data where new words in the form of slang or names emerge
over time. Another limitation is the fact that one has to state the
number of topics the model should use. This is not a big issue, but
could lead to some of the resulting topics to be insignificant.
\begin{center}
\begin{figure}[h]
\centering
\includegraphics[width=0.6\textwidth]{figs/LDA_standard_model.png}
\caption{Plate notation of the standard LDA model}
\end{figure}
\end{center}

@@

In LDA, our ultimate goal is to find the posterior distribution of
latent topic variables after observing training data. However,
computing this distribution is intractable, so we're forced to
approximate. One approximation approach is to use an optimization
method called Variational Bayes. In the original paper it's being used
a variational Bayes approximation of the posterior distribution;
\cite{blei2003latent} alternative inference techniques use Gibbs
sampling and expectation propagation \todo{Cite these techniques.
  AFAIK, if you do not use Variational Bayes, you use MCMC with Gibbs
  sampling.}. The goal of variational
inference is to approximate an intractable probability distribution
$p$, with a tractable one $q$, in a way that makes them as close as
possible given a similarity measure. In this thesis, variational
inference is used to approximate the posterior distribution of latent
variables. A new distribution over the hidden variables $q(z, \beta)$
(called the variational distribution) is introduced. This new
distribution has properties such that it can be efficiently
computed.\cite{Fox2011ATO}

@@

    \Repeat{change in $\phi < \epsilon$}{
        update $\phi$ 
        update $\gamma$ 
     }
}
M Step:

update $\lambda$
\end{algorithm}
\todo{Update $\phi$ and $\gamma$ do not show as separate lines.}

The specific updates are:
\begin{equation}
\phi_{d,word_{i},k} \propto e^{E_{q}(log \theta_{d,k})+E_{q}(log\beta_{k,word_{i}})}
\end{equation}
\begin{equation}
\gamma_{d,k} = \alpha + \sum_{word_{i}}\phi_{d,word_{i},k}n_{d,word_{i}}
\end{equation}
\begin{equation}
\lambda_{k,word_{i}}=\eta +\sum_{d}n_{d,word_{i}}\phi_{d,word_{i},k}
\end{equation}
Where $n_{d,word_{i}}$ is the number of occurrences of $word_{i}$ in
document $d$. The algorithm above has constant memory requirements and
empirically converges faster than collapsed Gibbs sampling. However,
it still requires a full pass through the entire corpus in each
iteration. It can be slow to apply to very large datasets.
\todo{Provide citation for the description you have given; I assume
  it's the same blog post as for the following paragraph.}

\paragraph{Online variational LDA}

An online variational inference algorithm is proposed by Hoffman et all.\cite{onlineLDAvb} for fitting $\lambda$, the parameters to the variational
posterior over the topic distributions $\beta$. The "online LDA" is nearly as
simple as the variational algorithm, but converges much faster for large
datasets. Variational inference requires to pass through the whole dataset for
each iteration, and that can be a problem when working with big datasets and is
not an option to use cases where the data is arriving constantly. The Online
Variational Inference algorithm aims to solve both problems without compromising
the quality of the obtained topics and with an algorithm nearly as simple as
Variational Inference. For this algorithm we will only highlight the differences
respect the Variational Inference and the resulting algorithm. The key
difference is the way to fit the variational parameter $\lambda$ respect the
Variational Inference explained in the previous section.

@@

\section{Methodology}

>>>>

Nowadays, there are some funding opportunities for researchers, but are 
really competitive. The European Union countries are encouraged to
invest 3\% of their GDP in research and development by 2020. EU's
latest research programme, Horizon 2020, allows researchers to
identify new opportunities and directions in almost any field of
research. The programme's goals are to strengthen the EU's position in
science, support industrial innovation and address social concerns,
such as climate change, green transportation and energy.

Another funding agency that actively promotes the progress of science
is the National Science Foundation in the United States of America.
NSF supports research in all the non medical fields of science \&
engineering and aims to advance the national health and prosperity. It
funds approximately 24\% of the federally supported research
conducted by the USA's colleges and universities.\footnote{\url{https://www.nsf.gov/about/}}

With our work, we aim to investigate the evolution of scientific
research in the EU and the US. In order to gain some insights  regarding
the research development, we used as a dataset the projects that got funded 
the last two decades under the EU Framework Programmes and the NSF in USA.

When researchers apply for funding, one of the first steps they do is
to identify an interesting subject of the many possible topics of
scientific investigation. The topic(s) of a project is always
described in the project's abstract/description. Using these abstracts, 
one can possibly investigate the topics of the projects and find the 
subjects that are being addressed and gain some information regarding 
the popularity of the topics among others.

In this work, we discover topics and their evolution from 1994 to
2017, using the projects' abstracts that got funded in this
period. Using a topic model, we extract a set of topics and
investigate the changes in the content per year/framework programme.
These topics are used to gain insight into some of the research that
occurs the last two decades. Our goal is to share these insights with
researchers and encourage them to develop better strategies to
establish themselves as more effective researchers globally, by
choosing a subject that is more likely to get a funding.

In section \ref{theoretic} introduces some basic concepts
in topic modeling and we describe the model that we used for topic
extraction, Latent Dirichlet Allocation, which is a statistical
generative model for documents. In section \ref{relatedwork}, 
we briefly talk about the papers that mostly influenced our study. 
In section \ref{methodology} we provide information
about the dataset structure and the methodology that we followed in
our analysis. In section \ref{analysis} we interpret the results of 
our analysis, and we discuss the evolution of the topics over time and the
differences between the two regions' topics based on their similarity.
Finally, section \ref{conclusion} summarizes our work and section 
\ref{future} suggests some future work.

\section{Theoretic background}
\label{theoretic}
In this chapter we will present some basic theoretic background about
topic modeling. To deal with extensive text body \todo{probably
  plural}, machine learning researchers have developed topic modeling:
a suite of methods that analyze the words of the original text to
discover the themes that run through the data and detect hidden
semantic structures by analyzing a corpus. By marking up the documents
using these topics and by using the resulting structure, the model can
be used for organization, finding similar documents, or any other
similar tasks. A topic is here defined as a probability distribution
of words, meaning that certain words are more likely within a certain
topic. Since the topics emerge from the analysis of the original
texts, topic modeling algorithms do not require any prior annotations
or labeling of the documents.\cite{Blei11introductionto}

@@

Some further common concepts and terms in topic modeling: 
\begin{itemize}
\item[] A \textit{word} is defined as an item from a vocabulary
  indexed from ${1, ..., V}$, where $V$ is the size of the
  vocabulary. All words are represented as unit-basis vectors with one
  component equal to one and the rest equal to zero.
\item[] A \textit{document} is a collection of words denoted by
  $w = {w_1, ..., w_N}$, where $w_n$ is the $n$th word and $N$ is the
  total number of words in the collection.
\item[] A \textit{corpus} is a collection of documents in a dataset.
  It is denoted $D = {w_1,..., w_M}$, where $w_m$ is the $m$th
  document in the corpus and $M$ is the total number of documents.
\item[] \textit{Latent variables} are variables that may not be directly observed, 
unlike observable variables. Latent variables can instead be inferred 
from other observable variables.
\item[] \textit{Polysemy} is the capacity for a word to have multiple related meanings. 
An example of this is the word “plant” which can mean a living organism of 
the kind exemplified by trees, herbs etc., or a place where an industrial or
manufacturing process takes place.
\item[] \textit{Synonymy} is the capacity for several words to have similar meanings such as the words “buy” and “purchase”.
\end{itemize}

@@

When introducing the Latent Dirichlet Allocation in section \ref{LDAexplained}, 
we will be using the terms prior and posterior probabilities
and likelihood function:

\begin{itemize}
\item The prior probability distribution of an uncertain
  quantity is the probability distribution that express the value of
  this quantity before some evidence is taken into account.
\item The posterior probability is the conditional probability after some 
evidence is taken into account.
\item The likelihood function of a set of parameters given some outcome 
is equal to the probability of observing said outcome given those 
parameter values.
\end{itemize}

@@

In the context of probabilities, the Jensen Inequality is stated as
follows: If $X$ is a random variable and $f$ a
convex function then:

@@

And, if $f$ is a concave function the inequality turns:

@@

where $\mathbb{E}$ is the expectation of the random variable $X$.
\footnote{Suppose random variable $X$ can take value $x_{1}$ with 
probability $p_{1}$, value $x_{2}$ with probability $p_{2}$, 
and so on, up to value $x_{k}$ with probability $p_{k}$. 
Then the expectation of this random variable $X$ is defined as
$\mathbb{E}[X] = x_{1}p_{1} + x_{2}p_{2} + ... + x_{k}p_{k}$.}

@@

\begin{equation}
D_{KL}(P \mid \mid Q) = \int_{-\infty}^{+\infty} p(x) \log\frac{p(x)}{q(x)}dx
\end{equation}

This can also be expressed using the definition of expectation (and
expressing the $\log$ of a division as difference of $\log$) as:

\begin{equation}
D_{KL}(P \mid \mid Q) = \mathbb{E}_{p}[\log{p(x)}] - \mathbb{E}_{p}[\log{q(x)}]
\end{equation}

@@

The probability distribution that the model uses is the multinomial, 
because it assumes that this is the distribution of topics in 
documents and words in topics (read Latent Dirichlet Allocation in section \ref{LDAexplained}). 
In probability theory, the multinomial distribution is a generalization
of the binomial distribution. For example, it models the probability
of counts for rolling a $k$-sided dice $n$ times. For $n$ independent trials
each of which leads to a success for exactly one of $k$ categories, with
each category having a given fixed success probability, the
multinomial distribution gives the probability of any particular
combination of numbers of successes for the various
categories\footnote{\url{https://en.wikipedia.org/wiki/Multinomial_distribution}}.

@@

The Dirichlet distribution is the conjugate prior of the multinomial
distribution. This means that if the prior distribution of the
multinomial parameters is Dirichlet then the posterior distribution of the
parameters is still a Dirichlet distribution.

@@

First, we'll introduce some basic models for topic modeling. The
Latent Semantic Indexing, or LSI, was presented by Deerwester et al.
in 1990.\cite{Deerwester90indexingby} The model manages to deal with the
problem that multiple terms can refer to the same meaning, i.e.
synonymy. However, it is not as successful regarding polysemy. The
reason is that every term is represented as just one point in the
so-called latent semantic space. Furthermore, a word that can mean two
or more different things is represented as a weighted average of the
different meanings.

After LSI was introduced, Hofmann presented the Probabilistic Latent
Semantic Indexing, or PLSI model\cite{Hofmann:1999:PLS:312624.312649}. 
PLSI is a topic model where each word in a document is generated 
from a single topic which results in that each document in a corpus 
can be represented with a topic distribution. Later, in 2003 Blei et al. 
presented the Latent Dirichlet Allocation, or LDA\cite{blei2003latent}. 
As opposed to PLSI, LDA is a statistically generative model for documents 
where each word in a document can be generated by all topics.

@@

Latent Semantic Indexing was one of the earliest methods for finding
relationships between documents and the words that occur in
them.\cite{Deerwester90indexingby} In LSI a term-document matrix is
created by analyzing the corpus, where the rows correspond to words
and columns to documents. Each element in this sparse matrix describes
the number of times a word occurs in a document but this term count
can also be weighted with for instance TF-IDF.\footnote{TF-IDF stands 
for term frequency - inverse document frequency, and the tf-idf weight is 
a weight often used in information retrieval and text mining. This weight 
is a statistical measure used to evaluate how important a word is to 
a document in a collection or corpus. The importance increases
proportionally to the number of times a word appears in the document (TF) but
is offset by the frequency of the word in the corpus (IDF).} Letting
each word represent a dimension in a very high dimensional space, a
document can be seen as a vector with components corresponding to its
weighted term counts. \cite{Salton:1988:TAA:54259.54260}

@@

\subsubsection{Latent Dirichlet Allocation (LDA)}
\label{LDAexplained}
LDA is a topic model that was first presented as a graphical model for topic
discovery by David Blei, Andrew Ng, and Michael I. Jordan in 2003. The basic
idea is that documents exhibit multiple topics. LDA is a generative statistical
model that allows sets of observations to be explained by unobserved groups
that explain why some parts of the data are similar. For example, if
observations are words collected into documents, it posits that each document
is a mixture of a small number of topics and that each word's occurrence is
attributable to one of the document's topics.\cite{Blei11introductionto}

A limitation with PLSI is that there is no generative model for these
proportions, making it difficult to handle unseen documents. The reduced
document description result in PLSI is a vector where each element describes the
mixing proportion for a topic. The Latent Dirichlet Allocation model tries to
solve this limitation by setting a Dirichlet prior on the topic
distribution.\cite{blei2003latent} In practice, this results in more reasonable
mixtures of topics in a document.

@@

The basic idea in LDA is that we define a topic to be a multinomial
distribution over a fixed vocabulary. Technically, the model assumes 
that the topics are generated first, before the
documents.\cite{Blei11introductionto} Now for each document in the
collection, we generate the words in a two-stage process:

@@

This statistical model reflects the intuition that documents exhibit
multiple topics. Each document exhibits the topics with different
proportion (step 1); each word in each document is drawn from one of
the topics (step 2b), where the selected topic is chosen from the
per-document distribution over topics (step 2a). All the documents in
the corpus share the same set of topics, but each document exhibits
those topics with different proportion - that's the distinguishing
characteristic of LDA. A more detailed explanation of the model
follows in section \ref{LDAmodel}.

\paragraph{Model}
\label{LDAmodel}
LDA is a probabilistic model of text documents, which assumes a
collection of K topics. Each topic defines a multinomial distribution
over the vocabulary and is assumed to have been drawn from a Dirichlet
$\beta_k \sim Dirichlet(\eta)$, where $beta_k$ is the posterior 
distribution of the topic $k$.

In LDA, each document is represented by a mixture of the topics, with
weight $\theta_w$ for topic
$z$ in document $w$. These weights have a Dirichlet
prior, parametrized by a hyper-parameter $\alpha$. Each topic is a
probability distribution over the vocabulary, with probability
$\beta^w_z$ for word $w$ in topic $z$. We define the following
variables and notation:
\begin{itemize}
  \item[] $k$ is the number of topics.
  \item[] $V$ is the number of unique words in the vocabulary.
  \item[] $\theta$  is the topic distribution (of length $k$) for a document,
   drawn from a uniform Dirichlet distribution with parameter $\alpha$.
 \item[] $z_{n}$ is a topic assignment for word $w_{n}$, 
 sampled from $p(z_{n} = i|\theta) = \theta_{i}$.
  \item[] \textbf{w}$ = (w_{1}, ... , w_{N})$ is a document with $N$ words.
  \item[] $w_{n}^{i} = 1$ means that the word $w_{n}$ is the $i$-th word 
  of the vocabulary.
  \item[] $\beta$  is a $k \times V$ matrix, where each row $\beta_{i}$ 
  is the multinomial distribution for the $i$-th topic. That is, 
  $\beta_{ij} = p(w^{j} = 1 | z_{j} = i)$.
\end{itemize}

@@

We can see that this does in fact generate a document based on the topic 
mixture $\theta$, the topic-word assignments $z$, and the probability 
matrix $\beta$. Then, we can analyze a corpus of documents with LDA by 
examining $\beta$, $\theta$ 
and $z$ conditioned on the documents. This reveals latent structure in
the collection that can be used for prediction or data exploration.

@@

This
posterior can't be computed directly; it is usually approximated using Markov
Chain Monte Carlo (MCMC) or variational inference. Both methods are effective,
but also present significant computational challenges in large data sets
~\cite{onlineLDAvb}. In our implementation, we're using the online variational
LDA model, which is described in section \ref{oLDAvb}.

A limitation with the standard LDA model is the fixed 
vocabulary. This means that new words cannot enter the model and that
you have to know a priori what words will model the corpus
effectively. A setting where this could be problematic is when
analyzing data where new words in the form of slang or names emerge
over time. Another limitation is the fact that one has to state the
number of topics the model should use. This is not a big issue, but
could lead to some of the resulting topics to be insignificant.
\begin{center}
\begin{figure}[h]
\centering
\includegraphics[width=0.6\textwidth]{figs/LDA_standard_model.png}
\caption{Plate notation of the standard LDA model}
\end{figure}
\end{center}

@@

In LDA, our ultimate goal is to find the posterior distribution of
latent topic variables after observing training data. However,
computing this distribution is intractable, so we're forced to
approximate. One approximation approach is to use an optimization
method called Variational Bayes. In the original paper it's being used
a variational Bayes approximation of the posterior distribution;
\cite{blei2003latent} alternative inference techniques use 
MCMC with Gibbs sampling\cite{Griffiths02gibbssampling} and expectation 
propagation.\cite{Minka:2002:EGA:2073876.2073918} The goal of variational
inference is to approximate an intractable probability distribution
$p$, with a tractable one $q$, in a way that makes them as close as
possible given a similarity measure. In this thesis, variational
inference is used to approximate the posterior distribution of latent
variables. A new distribution over the hidden variables $q(z, \beta)$
(called the variational distribution) is introduced. This new
distribution has properties such that it can be efficiently
computed.\cite{Fox2011ATO}

@@

        update $\gamma$ 
     }
}
M Step:

The specific updates are:
\begin{equation}
\phi_{d,word_{i},k} \propto e^{E_{q}(log \theta_{d,k})+E_{q}(log\beta_{k,word_{i}})}
\end{equation}
\begin{equation}
\gamma_{d,k} = \alpha + \sum_{word_{i}}\phi_{d,word_{i},k}n_{d,word_{i}}
\end{equation}
\begin{equation}
\lambda_{k,word_{i}}=\eta +\sum_{d}n_{d,word_{i}}\phi_{d,word_{i},k}
\end{equation}
Where $n_{d,word_{i}}$ is the number of occurrences of $word_{i}$ in
document $d$. The algorithm above has constant memory requirements and
empirically converges faster than collapsed Gibbs sampling. However,
it still requires a full pass through the entire corpus in each
iteration. It can be slow to apply to very large datasets.
\footnote{\url{https://wellecks.wordpress.com/2014/10/26/ldaoverflow-with-online-lda/}}

\paragraph{Online variational LDA}
\label{oLDAvb}
An online variational inference algorithm is proposed by Hoffman et all.\cite{onlineLDAvb} for fitting $\lambda$, the parameters to the variational
posterior over the topic distributions $\beta$. The "online LDA" is nearly as
simple as the variational algorithm, but converges much faster for large
datasets. Variational inference requires to pass through the whole dataset for
each iteration, and that can be a problem when working with big datasets and is
not an option to use cases where the data is arriving constantly. The Online
Variational Inference algorithm aims to solve both problems without compromising
the quality of the obtained topics and with an algorithm nearly as simple as
Variational Inference. For this algorithm we will only highlight the differences
respect the Variational Inference and the resulting algorithm. The key
difference is the way to fit the variational parameter $\lambda$ respect the
Variational Inference explained in the previous section.

@@

\section{Related work}
\label{relatedwork}
This section surveys previous work in topic modeling on
scientific data. Topic modeling algorithms provide a
suite of algorithms to discover hidden thematic structures 
in large collections of texts and can be adapted to
many kinds of data. These algorithms can be used to summarize,
visualize and explore a collection of documents. The papers listed
below are applying a Latent Dirichlet Allocation model in an
interesting approach on their corpus.
